Agile Monitoring of Evolving Web Services

### 2018-03-21 restructured.tex
restructuring a bit...

%
\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services

@@

\section{Case Study}
\label{sec:case}

@@

\section{The Flask Dashboard}
\label{sec:tool}

%\todo{Refer to VISSOFT paper~\cite{vogel2017low} and rephrase where possible; add more info on implementation, e.g. annotation mechanism, database design, session handling etc.}

@@

  \paragraph{Usage \& Configuration}

  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive. \ml{this is not what matters... rather, what matters is that with the right backend, we can do this minimal configuration for any technology... do you see what i mean? let me see if i can put it in words...}

  Our main goal with the \tool was to allow the integration of the dashboard with an API with minimal effort.
  To start using the \tool, one simply needs to add two lines of code to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g. Django) by simply providing a few back-end adapters in the right places. }}:

@@

  After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application. Further configuration is possible by adding additional statements before the binding definition, for example,

  During binding, the \tool will search for all endpoints defined in the target application. These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}. The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.

    % \mltp{a small screenshot of how the dashboard allows one to select the interesting }
    \begin{figure}
      \centering
      \includegraphics[width=\linewidth]{selecting_endpoints.png}
      \caption{Endpoints that are available for monitoring in the \zee API}
      \label{fig:sep}
    \end{figure}

  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint. This way, the wrapper will be executed whenever that API call is made. The wrapper contains the code that takes care of monitoring an endpoint. Data collected by the wrappers are persisted in a local database. The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.

@@

  \paragraph{Automated outlier detection and monitoring}

  \ins{When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) 
  of particular interest to the API developers are performance {\em outliers}. Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application. Thus,} identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.

  \paragraph{User-level monitoring}

  \ins{Sometimes, the same service endpoint will have different execution times for different users, based on the amount of data that a user has in their account. Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \footnote{https://techcrunch.com/2010/03/14/gmail-slow/, visited September 23 2017}. }

  \ins{If the API provider needs to understand the performance of their service for individual users, or for different user groups, the dashboard should have a way of distilling a user or a user group from the API request. For a large number of Flask applications that use the global \code{flask.request} object, the object} can be used to retrieve session information, \ins{as in the following snippet}: 
  % . This ability can be used in order to allow, for example, the identification of application users from the \tool, and the consequent aggregation of monitoring information on a per user basis as follows:

  \begin{lstlisting}[style=custompython]  
  # app specific way of extracting the user
  # from a flask request object    
  def get_user_id():
  sid = int(flask.request.args['session'])
  session = User.find_for_session(sid)
  return user_id

  \end{lstlisting}
  This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage. It also provides a view on the distribution of the work load across the application endpoints on a per user basis that allows the identification of more or less popular endpoints for further performance improvements. \ml{this previous paragraph seems a bit reduntant. must revisit}

  \paragraph{Evolution monitoring for Git-versioned projects }

  Version control in \tool is supported in two ways.

  \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:

  \item \textit{Automatic} version control, aimed for use in conjunction with a version control system and currently only supporting Git\footnote{\url{https://git-scm.com/}}, is enabled by linking to the Git folder in the application source code:

      \begin{lstlisting}[style=custompython]
      ....
      dashboard.config.git = '.git'
      ....
      \end{lstlisting}

    With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started. Information about the version history of the application is also persisted in the tool database and is used during the collection of measurements for the application.

  \paragraph{Preemptive monitoring}

@@

\begin{figure*}[t]
	\centering
	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
	\qquad
	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
	\caption{Some of the available views\label{fig:views}}
\end{figure*}

@@

  \todo{remove the part about how to configure git, the discussion has been subsumed by Section~\ref{sec:tool}.}

  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization. One way to do this is to add an extra line of configuration to allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service:

%  \begin{lstlisting}[caption=Configuring the \tool with the path to the .git folder enables the generation of evolutionary performance graphs, style=custompython]
%
%dashboard.config.git = 'path/to/.git'
%
%
%    \end{lstlisting}

@@

  With this extra configuration, the \tool can now automatically detect the current version of the project, and group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}. \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom

%    \begin{figure}[h!]
%      \centering
%      \includegraphics[width=0.95\linewidth]{translation_endpoint_evolution_}
%      \caption{Visualizing The Performance Evolution of the \epTranslations endpoint}
%      \label{fig:tee}
%    \end{figure}

@@

\paragraph{User Experience}
\label{sec:user}

@@

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.85\linewidth]{time_per_user}
	\caption{The \epFeedItems shows a very high variability across users}
	\label{fig:tpu}
\end{figure}

@@

The limitation of the previous view is that it does not present the information also on a per version basis. To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined. Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle. The colors represent users. The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e. number of sources to which the user is registered) is the reason for the variation in response times.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.9\linewidth]{time_per_user_per_version}
	\caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
	\label{fig:tuv}
\end{figure}

@@

\subsection{Effectiveness \& Ease of Use}

\todo{@M: move here the war stories, and discuss how useful the whole thing was to the application maintainer}

\subsection{Efficiency of Preemptive Monitoring}

\begin{figure}
	\centering
	\subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
	\quad
	\subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
	\caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
	\label{fig:preemptive}
\end{figure}

\subsection{Discussion}

@@

@@

>>>>

%
\title{Agile Monitoring of Evolving Web Services}

@@

\section{Case Study: The API (Zeeguu)}
\label{sec:case}

@@

\newpage

\section{Case Study: The Technology (Flask)}
\label{sec:tool}

 Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
 \vspace{2cm}

 Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.

 “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality. The “micro” in microframework means Flask aims to keep the core simple but extensible. Flask won’t make many decisions for you, such as what database to use. Those decisions that it does make, such as what templating engine to use, are easy to change. Everything else is up to you, so that Flask can be everything you need and nothing you don’t.

 Flask has many configuration values, with sensible defaults, and a few conventions when getting started. By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively. While this can be changed, you usually don’t have to, especially when getting started.

 A Python web application based on WSGI has to have one central callable object that implements the actual application. In Flask this is an instance of the Flask class. Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?

from flask import Flask
app = Flask(__name__)

@app.route('/')
def index():
    return 'Hello World!'

\subsection*{NOTHING PREVENTS GENERALIZATION}
In this article we take as an example the Flask case study.
But there is nothing that prevents the generalization of 
our approach.

MAKE SURE THAT THIS IS CLEAR.

    \begin{figure*}[ht!]
      \centering
      \includegraphics[width=0.8\linewidth]{interceptor}
      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
      \label{fig:sep}
    \end{figure*}

\newpage

\section{Monitoring Service Utilization}

REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.

\ins{The first important question that such a 
service monitoring system should provide is 
information about what endpoints are being
used and by whom}.

TODO: Find references, arguments, related work
that provide more details about why understanding
who uses endpoints, and which are used are important.

This is clearly the case in static analysis (see paper
by haenni and lungu...) but should even more be the
case in APIs.

  \newpage
TODO: TALK ABOUT THE DATABASE...

@@

  \newpage
  \section{One Line Of Code Configuration With Technology Specific Interceptors}

  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.

  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }

  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g. Django) by simply providing a few back-end adapters in the right places. }}:

@@

  \ins{The dashboard can take advantage of the fact that it is monitoring an API... so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.

  During binding, the \tool will search for all endpoints defined in the target application. \ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance. In our case, the overhead here was quite small: TO MEASURE}.

  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint. This way, the wrapper will be executed whenever that API call is made before the actual function is called. The wrapper contains the code that takes care of monitoring an endpoint. Data collected by the wrappers are persisted in a local database. The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}

  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.

    \begin{figure}[h!]
      \centering
      \includegraphics[width=\linewidth]{selecting_endpoints.png}
      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
      \label{fig:sep}
    \end{figure}

  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).

  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.

@@

\begin{figure*}[h!]
  \centering
  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.7\columnwidth]{number_of_requests_}\label{fig:aeu}}
  \qquad
  \subfloat[requests per hour heatmap]{\includegraphics[width=0.8\columnwidth]{daily_patterns_}\label{fig:dp}}
  \subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.7\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
  \caption{Some of the available views\label{fig:views}}
\end{figure*}

\newpage
\section{How Far Can We Go With One Line of Configuration?}

A discussion must be added here about the types of useful information and problems that can be solved with these views.

@@

REQUIREMENT: Adding a time interval is important. 
Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.

@@

      \begin{figure*}[h!]
    \centering
    \includegraphics[width=\linewidth]{outliers}
    \caption{Automatically collected outlier information}
    \label{fig:figure1}
  \end{figure*}

  \newpage

  \section{Automated outlier detection and monitoring}

  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
  of particular interest to the API developers are performance {\em outliers}.

  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application. Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.

The bottom panel shows the stack trace. 
In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.

This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment. Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.

\subsection {A Special Type of Outlier: Exceptions}

   TODO: probably something for the journal extension ...
  statistics about which endpoints fail most often might be useful too.
  same information as the outlier maybe?

\begin{figure*}[h!]
  \centering
  \includegraphics[width=0.6\linewidth]{translation_endpoint_evolution_}
  \caption{The Performance Evolution of the \epTranslations endpoint}
  \label{fig:tee}
\end{figure*}

  \newpage
  \section{Evolution monitoring }

  When thinking about performance, 
  one can not avoid thinking about the performance evolution.

  Version control in \tool can be supported in two ways,
  the developer explicitly states the current version, 
  or the current version is automatically detected based
  on some version control system.

  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.

  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:

@@

  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}. \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom

@@

  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.

\todo{TODO: endpoint evolution }

\begin{figure*}[!ht]
  \centering
  \includegraphics[width=0.4\linewidth]{time_per_user}
  \caption{The \epFeedItems shows a very high variability across users}
  \label{fig:tpu}
\end{figure*}

\newpage
\section{Grouping Requests}
\label{sec:user}

@@

The limitation of the previous view is that it does not present the information also on a per version basis. To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined. Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.

The colors represent users. The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e. number of sources to which the user is registered) is the reason for the variation in response times. \ins{one can see that for user 1 performance degrades over the }

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
  \label{fig:tuv}
\end{figure}

  \newpage
  \section{Preemptive monitoring}

  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
  understanding the impact of software evolution on performance.
  However, just by observing this one can't be sure whether 
  performance changes because of the code or because of the payload.}

\begin{figure}[h!]
  \centering
  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
  \quad
  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
  \label{fig:preemptive}
\end{figure}

% \begin{figure*}[t]
% 	\centering
% 	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
% 	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
% 	\qquad
% 	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
% 	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
% 	\caption{Some of the available views\label{fig:views}}
% \end{figure*}

@@

@@

@@

\section{Discussion}

@@

  \subsection{Distributed System Monitoring}
  \todo{To think about}
  Supporting situations where the API is deployed across multiple containers for example.

@@

\newpage
\appendix

  \subsection{Changing the Default /dashboard Endpoint }

  THIS STUFF SHOULD BE MOVED TO THE APPENDIX. And in a footnote we just mention that ... \ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed. A way of overwriting the common sense defaults.

  Further configuration is not needed possible by adding additional statements before the binding definition, for example,

  \subsection{For those who don't use GIT}

    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
